---
title: "Cyclistic Biketrip data 2023"
author: "Mario"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---


# The dataset is comprised of six comma-separated value (CSV) files that represent the first half of 2023’s bike rides.
# The purpose of this study is to analyze trip data to develop strategies oriented to membership conversion.


## Load libraries into session
```{r}
library(tidyverse)
library(tidyr)
library(lubridate)
library(DescTools)
```


## Load and merge the data into a single dataset
```{r}
df1 <- read.csv("202301-divvy-tripdata.csv", na.strings=c("","NA"))
df2 <- read.csv("202302-divvy-tripdata.csv", na.strings=c("","NA"))
df3 <- read.csv("202303-divvy-tripdata.csv", na.strings=c("","NA"))
df4 <- read.csv("202304-divvy-tripdata.csv", na.strings=c("","NA"))
df5 <- read.csv("202305-divvy-tripdata.csv", na.strings=c("","NA"))
df6 <- read.csv("202306-divvy-tripdata.csv", na.strings=c("","NA"))
biketrip_df <- rbind(df1, df2, df3, df4, df5, df6)
tibble(biketrip_df)
```


## Remove observations 
```{r}
biketrip_clean <- biketrip_df %>% 
  drop_na()
```


## Reformat the “started_at” and “ended_at” variables to date-time objects
```{r}
biketrip_clean$started_at <- ymd_hms(biketrip_clean$started_at)
biketrip_clean$ended_at <- ymd_hms(biketrip_clean$ended_at)
```


## Dates
### Extract the date
```{r}
biketrip_clean$start_date <- as.Date(biketrip_clean$started_at)
```


### Extracting day
```{r}
biketrip_clean$day <- day(biketrip_clean$started_at)
```


### Extracting month
```{r}
biketrip_clean$month <- month(biketrip_clean$started_at)
```


### Extracting year
```{r}
biketrip_clean$year <- year(biketrip_clean$started_at)
```


### Calculate day of the week rides began
```{r}
biketrip_clean$day_of_week <- weekdays(biketrip_clean$started_at)
```


### Calculate trip duration
```{r}
biketrip_clean <- biketrip_clean %>%
  mutate(ride_length = ended_at - started_at)
```


## Drop negative trip duration
```{r}
biketrip_clean <- biketrip_clean %>%   
	filter(ride_length > 0)
```


## Drop columns that wont be used
```{r}
biketrip_clean = subset(biketrip_clean, select = -c(ride_id, start_station_id, end_station_id, start_lat, start_lng, end_lat, end_lng))
```


## Export cleaned df to a new csv
```{r}
write_csv(biketrip_clean, "2023-divvy-tripdata_final.csv")
biketrip_final <- read_csv("2023-divvy-tripdata_final.csv")
```


## Analysis

### Descriptive analysis of main variables
```{r}
analysis_df <- biketrip_final %>%
  group_by(member_casual) %>%
  summarize(
    mean_ride_length = mean(ride_length),
    max_ride_length = max(ride_length),
    mode_day_of_week = Mode(day_of_week),
    count_of_trips = n())
write_csv(analysis_df, "2023-divvy-analysis_df.csv")
```


### Total number of trips by biketype
```{r}
biketype_trip_count <- biketrip_final %>%
  group_by(member_casual, rideable_type) %>% 
  summarize(count_of_trips = n())
write_csv(biketype_trip_count, "2023-divvy-biketype_trip_count.csv")
```


### Station summary: member/casual
```{r}
station_analysis <- biketrip_final %>% 
  mutate(station = start_station_name) %>%
  group_by(start_station_name, member_casual) %>%  
  summarize(count_of_trips = n()) %>%    
  arrange(desc(count_of_trips))
write_csv(station_analysis, "2023-divvy-station_analysis.csv")
```


### Ride length: month
```{r}
ridelength_month <- biketrip_final %>%
  mutate(month_name = month.name[month]) %>% 
  mutate(month_name = factor(month_name, levels = c("January", "February", "March", "April", "May", "June"))) %>%
  group_by(member_casual, month_name) %>%
  summarize(
    mean_ride_length = mean(ride_length),
    count_of_trips = n()) %>% 
  mutate(mean_minutes = minute(seconds_to_period(mean_ride_length)))
write_csv(ridelength_month, "2023-divvy-ridelength_month.csv")
```


### Ride length: day of week
```{r}
ridelength_day_of_week <- biketrip_final %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarize(
    mean_ride_length = mean(ride_length),
    count_of_trips = n()) %>% 
  mutate(mean_minutes = minute(seconds_to_period(mean_ride_length)))
write_csv(ridelength_day_of_week, "2023-divvy-ridelength_day_of_week.csv")
```


# The following visualizations were not used. Instead, they were later replaced with Tableau visualizations.


### Visualize the number of rides by type of rider grouped by weeks
```{r}
ridelength_day_of_week %>% 
  arrange(day_of_week) %>% 
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  ggplot(aes(x = day_of_week, y = count_of_trips, color = member_casual, group = member_casual)) +
  geom_point() + 
  geom_line()
```


### Visualize the totalnumber of rides by type of rider grouped by month
```{r}
ridelength_month %>% 
  arrange(month_name) %>% 
  mutate(month_name = factor(month_name, levels = c("January", "February", "March", "April", "May", "June"))) %>%
  ggplot(aes(x = month_name, y = count_of_trips, group = member_casual, color = member_casual)) +
  geom_point() +
  geom_line()
```